Clines from PDE

### 2015-11-03 notes/pde-clines.Rmd
c code?

```{r doc_setup, include=FALSE}
fig.dim <- 5
knitr::opts_chunk$set(fig.width=2*fig.dim,fig.height=fig.dim,fig.align='center')
library(ReacTran)
source("tran.1D.R") # fix for log.A
```

@@

The PDE above gives probabilities given the initial background,
so therefore *linkage* to the selected site.
For the cline shape we need to average over choice of background,
which we denote
$$\begin{aligned}
p(x,\tau,r) =  p(x,\tau) q(x,A,\tau,r) + (1-p(x,\tau))q(x,B,\tau,r) .
\end{aligned}$$
Note that
$$\begin{aligned}
q(x,A,\tau,r) &= 1-q(-x,B,\tau,r) \\
p(x,\tau) &= 1-p(-x,\tau) 
\end{aligned}$$
and so
$$\begin{aligned}
p(x,\tau,r) =  p(x,\tau) q(x,A,\tau,r) + p(-x,\tau) (1-q(-x,A,\tau,r))A .
\end{aligned}$$

@@

The smaller $r$ is, the slower the per-background probabilities smooth out:
```{r small_r_sim}
r <- 0.01
pde.soln <- solve_pde( u=log_u, um1=log_1mu, r=r, times=tt, 
                  s=s, sigma=sigma, grid=thegrid, log.u=TRUE )
image(pde.soln,legend=TRUE,main=sprintf("q for z=%s, r=%0.3f",c("A","B"),r), yaxt='n', ylab='space',zlim=c(0,1))
axis(2,at=(pretty(thegrid$x.mid)-min(thegrid$x.mid))/diff(range(thegrid$x.mid)),labels=pretty(thegrid$x.mid))
pde.cline <- cline_from_soln(pde.soln,grid=thegrid)
layout(1)
clineplot(pde.cline,x=thegrid$x.mid,main=sprintf("clines, at r=%0.2f, up to t=%d",r,zone.age))
```

And, the cline flattens out *very* slowly:
```{r longer_t_sim}
r <- 1.0
zone.age <- 2000  # this is tau
tt <- seq(0,zone.age,length.out=100)
pde.soln <- solve_pde( u=log_u, um1=log_1mu, r=r, times=tt, 
                  s=s, sigma=sigma, grid=thegrid, log.u=TRUE )
image(pde.soln,legend=TRUE,main=sprintf("q for z=%s, r=%0.3f",c("A","B"),r), yaxt='n', ylab='space',zlim=c(0,1))
axis(2,at=(pretty(thegrid$x.mid)-min(thegrid$x.mid))/diff(range(thegrid$x.mid)),labels=pretty(thegrid$x.mid))
pde.cline <- cline_from_soln(pde.soln,grid=thegrid)
layout(1)
clineplot(pde.cline,x=thegrid$x.mid,main=sprintf("clines, at r=%0.2f, up to t=%d",r,zone.age))
```

@@

There are various haplotype-related statistics we could compute:

1.  Distance to closest recombination in a heterozygote

@@

The first haplotype-related statistic
is the distance along the chromosome from an $A$ allele until the closest recombination
that occurred in an $AB$ heterozygote.
Let $H$ be this distance;
then, 
$$\begin{aligned}
    u(x,\tau,r)=\P^(x,\tau)\{H>r\}
\end{aligned}$$
solves the following PDE:
$$\begin{aligned}
    u(x,0,r) &= 1  \\
    \partial_t u(x,t,r)
        &= 
        \grad \log(p(x)) \cdot \grad u(x,t,r) + \frac{\sigma^2}{2} \Delta u(x,t,r) - r (1-p(x)) u(x,t,r) .
\end{aligned}$$

>>>>

```{r doc_setup, include=FALSE}
fig.dim <- 5
knitr::opts_chunk$set(fig.width=2*fig.dim,fig.height=fig.dim,fig.align='center')
library(ReacTran)
library(Matrix)
source("tran.1D.R") # fix for log.A
```

@@

The PDE above gives probabilities given the initial background,
so therefore *linkage* to the selected site.
For the cline shape we need to average over choice of background,
which we denote
$$\begin{aligned}
p(x,\tau,r) =  p(x,\tau) q(x,A,\tau,r) + (1-p(x,\tau))q(x,B,\tau,r) .
\end{aligned}$$
Note that
$$\begin{aligned}
q(x,A,\tau,r) &= 1-q(-x,B,\tau,r) \\
p(x,\tau) &= 1-p(-x,\tau) 
\end{aligned}$$
and so
$$\begin{aligned}
p(x,\tau,r) =  p(x,\tau) q(x,A,\tau,r) + p(-x,\tau) (1-q(-x,A,\tau,r)) .
\end{aligned}$$

@@

The smaller $r$ is, the slower the per-background probabilities smooth out:
```{r small_r_sim,cache=TRUE}
r <- 0.01
pde.soln <- solve_pde( u=log_u, um1=log_1mu, r=r, times=tt, 
                  s=s, sigma=sigma, grid=thegrid, log.u=TRUE )
image(pde.soln,legend=TRUE,main=sprintf("q for z=%s, r=%0.3f",c("A","B"),r), yaxt='n', ylab='space',zlim=c(0,1))
axis(2,at=(pretty(thegrid$x.mid)-min(thegrid$x.mid))/diff(range(thegrid$x.mid)),labels=pretty(thegrid$x.mid))
pde.cline <- cline_from_soln(pde.soln,grid=thegrid)
layout(1)
clineplot(pde.cline,x=thegrid$x.mid,main=sprintf("clines, at r=%0.2f, up to t=%d",r,zone.age))
```

And, the cline flattens out *very* slowly:
```{r longer_t_sim,cache=TRUE}
r <- 1.0
zone.age <- 2000  # this is tau
tt <- seq(0,zone.age,length.out=100)
pde.soln <- solve_pde( u=log_u, um1=log_1mu, r=r, times=tt, 
                  s=s, sigma=sigma, grid=thegrid, log.u=TRUE )
image(pde.soln,legend=TRUE,main=sprintf("q for z=%s, r=%0.3f",c("A","B"),r), yaxt='n', ylab='space',zlim=c(0,1))
axis(2,at=(pretty(thegrid$x.mid)-min(thegrid$x.mid))/diff(range(thegrid$x.mid)),labels=pretty(thegrid$x.mid))
pde.cline <- cline_from_soln(pde.soln,grid=thegrid)
layout(1)
clineplot(pde.cline,x=thegrid$x.mid,main=sprintf("clines, at r=%0.2f, up to t=%d",r,zone.age))
```

@@

Consider a genomic region $[a,b]$, with $a\le 0\le b$,
with the selected site at 0.
Recombination without coalescence makes the history of this region
a branching process with conserved branching rates.
Each branch is labeled by a region of the genome,
and moves independently according to generator $G_r$, 
where $r$ is the distance from the region to the selected site,
and branches at rate $w$, where $w$ is its width.
We would like to know the probability that 
*all* branches end up on the same side of the barrier at time $\tau$ in the past.

Suppose that the region includes a type $A$ allele,
and we want the probability that it inherits entirely from the $A$ side.
Let
$$\begin{aligned}
    g(x,A,a,b,\tau) = \P^{x,A}\{ [a,b] \text{ inherits from $A$, $\tau$ generations ago }\} .
\end{aligned}$$
If $W(t)=\{W_1(t),W_2(t),\ldots,W_{n(t)}(t)\}$ is the state of the process $t$ generations ago,
where $W_k=(x_k,z_k,a_k,b_k)$,
with $x_k$ the position and $W_k\in\{A,B\}$ the allele,
then by independence of branches,
$$\begin{aligned}
    g(x,A,a,b,0) &= 1 \\
    g(x,B,a,b,0) &= 0 \\
    g(x,B,a,b,\tau) &= 0 \qquad \text{if } 0 \in [a,b] \\
    g(x,A,a,b,\tau)
    &=
    \E\left[
        \prod_{i=1}^{n(t)} g(x_k,z_k,a_k,b_k,\tau-t)
    \right] \qquad \text{for each } t \le \tau .
\end{aligned}$$
Taking $t = \epsilon$ in the last equation,
so that the probability of branching is $1-e^{-(b-a)\epsilon}=\epsilon(b-a)+O(\epsilon^2)$,
if $0\le a < b$ this is
$$\begin{aligned}
g(x,A,a,b,\tau)
    &=
    \epsilon 
    \left\{
    G_{r(a,b)} g(x,A,a,b,\tau)
    +
     \int_a^b g(x,A,a,\theta,\tau-\epsilon)
     \left(
     p(x,\tau-\epsilon)g(x,A,\theta,b,\tau-\epsilon) 
     + (1-p(x,\tau-\epsilon))g(x,B,\theta,b,\tau-\epsilon) 
     \right)d\theta
    \right\}
    + O(\epsilon^2),
\end{aligned}$$
where 
$$\begin{aligned}
r(a,b)=\begin{cases}
    \min(|a|,|b|) \qquad &\text{if } ab \ge 0 \\
    0 \qquad &\text{if } 0 \in [a,b] .
\end{cases}
\end{aligned}$$
and $G_r f(x,z)$ is defined above.
On the other hand, if $a<0<b$, then
$$\begin{aligned}
g(x,A,a,b,\tau)
    &=
    \epsilon 
    \left\{
    G_{r(a,b)} g(x,A,a,b,\tau)
    +
     \int_0^b g(x,A,a,\theta,\tau-\epsilon)
     \left(
     p(x,\tau-\epsilon)g(x,A,\theta,b,\tau-\epsilon)
     + (1-p(x,\tau-\epsilon))g(x,B,\theta,b,\tau-\epsilon) 
     \right)d\theta
    +
     \int_a^0 \left(
     p(x,\tau-\epsilon)g(x,A,a,\theta,\tau-\epsilon)
     + (1-p(x,\tau-\epsilon))g(x,B,a,\theta,\tau-\epsilon) 
     \right) g(x,A,\theta,b,\tau-\epsilon) d\theta
    \right\}
    + O(\epsilon^2),
\end{aligned}$$

Therefore, assuming $G$ is sufficiently nice,
$$\begin{aligned}
\frac{d}{d\tau} g(x,z,a,b,\tau)
&=
G_{r(a,b)} g(x,z,a,b,\tau)
+
\int_a^b
\left(
    g(x,z,a,\theta,\tau)g(x,z,\theta,b,\tau) 
    - g(x,z,a,b,\tau)
\right)
d\theta
\end{aligned}$$
Since $z \in \{A,B\}$, this is a system of two integro-differential equations 
in three variables plus time ($x$, $a$, $b$, and $\tau$; four if $x\in\R^2$).

Numerical solutions will discretize; what discretization of the genome might be good?
If we want to solve out to time $\tau$,
then on a chromosome of length 1 there will be about $\tau$ breakpoints,
each of which approximately Exponential($\tau$),
and so the smallest distance will have mean $1/\tau^2$.
But, suppose we've solved it for time $\tau$, and want to solve it to time $2\tau$.
We can use the same discretization, with initial conditions from the first round.
<!--
Quick check:
```{r smallest_check}
f <- function (t) { min(diff(c(0,sort(runif(rpois(1,lambda=t))),1))) }
fk <- sapply( 2:200, function (k) { replicate(n=100,f(k)) } )
plot(1+col(fk),fk,col=rainbow(300)[col(fk)],log='y',ylab='smallest interval')
points(2:200,colMeans(fk),pch=20,cex=2)
lines(2:200,1/(2:200)^2,col='purple',lwd=2)
```
-->

Note that if we think of $u(a,b)$ as a kernel, 
then the map $u(a,b) \mapsto \int_a^b u(a,\theta) u(\theta,b) d\theta$
is just the product of the kernel to itself;
on discretization this becomes a matrix power,
so the operator $u(a,b) \mapsto \int_a^b u(a,\theta)u(\theta,b) - u(a,b) d\theta$
```{r solve_branching}
xgrid <- setup.grid.1D(x.up=-20, x.down=20, N=100)
agrid <- setup.grid.1D(x.up=-0.5, x.down=0.5, N=30)
# determines the order the solution is in:
xab <- subset( expand.grid( x=xgrid$x.mid, a=agrid$x.mid, b=agrid$x.mid ), ( b>0 ) & (abs(a)<=abs(b)) )
find_ab <- function (ab) {
    # map into the subset of (a,b) contained in xab
    flip <- ab[,1]>ab[,2] 
    ab[flip,] <- ab[flip,] %*% cbind( c(0,1), c(1,0) )
    flop <- ( ab[,1]<0 & abs(ab[,1])>=abs(ab[,2]) )
    ab[flop,] <- ab[flop,] %*% cbind( c(0,-1), c(-1,0) )
    return(ab)
}
integral_map <- function (y,ab) {
    # the discretization of the map y(a,b) -> \int_a^b (h(a,u)*h(u,b)-h(a,b)) du
    # The key step is to do something like:
    #     1 | x y .
    #   a 2 | . . u
    #     3 | . . v
    #       +------
    #     b   1 2 3
    # --> should  return x*u + y*v,
    # the inner product of a partial row with a partial column

}
```

@@

A simpler statistic, that might approximate the correct quantity well,
is the distance along the chromosome from an $A$ allele until the closest recombination
that occurred in an $AB$ heterozygote.
Let $H$ be this distance;
then, 
$$\begin{aligned}
    u(x,\tau,r)=\P^{(x,\tau)}\{H>r\}
\end{aligned}$$
solves the following PDE:
$$\begin{aligned}
    u(x,0,r) &= 1  \\
    \partial_t u(x,t,r)
        &= 
        \grad \log(p(x)) \cdot \grad u(x,t,r) + \frac{\sigma^2}{2} \Delta u(x,t,r) - r (1-p(x)) u(x,t,r) .
\end{aligned}$$
